CrystalPlan: an Experiment Planning Tool for Crystallography

### 2011-02-11 docs/CrystalPlan_paper/paper.tex
Fixes #182: Final revisions for CrystalPlan paper, to be published in J Appl Cryst

\newif\ifdograph
\dographtrue % DO keep the graphs.
\dographfalse % DONT put graphs - makes things faster

@@

\begin{abstract}
Beam time at large user program based x-ray and neutron scattering facilities is
in high demand and always at a premium. CrystalPlan, a highly efficient
experiment planning software has been developed to maximize the use of available
beamtime per sample per experiment. 
This program can calculate and optimize the data coverage of a crystal in
reciprocal space in a single-crystal diffraction time-of-flight experiment.
CrystalPlan can help a user build an experiment plan that will acquire the most
data possible, with sufficient coverage but limited redundancy, therefore
increasing scientific productivity. 
A user friendly GUI (Graphical User Interface) including a 3D viewer, an
automated coverage optimizer, and an option to reorient the crystal for the measurement 
of selected $hkl$s on specific detector positions 
are among its useful features. 
A sample use case of the program with the TOPAZ beamline at SNS will be
presented. 
\end{abstract}

@@

The TOPAZ beamline at the Spallation Neutron Source (SNS) at Oak Ridge National
Laboratory (ORNL), currently finishing commissioning, is a neutron
time-of-flight Laue diffractometer with high wavelength (time) resolution. It
is designed to acquire single-crystal diffraction data of small-to-moderate sized unit cells with  high
throughput. The final design calls for an array of 48 detectors covering three
quarters of $2\pi$ steradians. However as of this writing, only 14 of the 48
detectors are installed, giving only a quarter of $2\pi$ steradian coverage. 
To have sufficient data for a reliable structure solution and
refinement, an experimenter will typically want to measure $ > 85\%$ of the
peaks within a certain range of d-spacings. However, TOPAZ's wide wavelength
band combined with its complex geometry makes it difficult to come up with
an effective set of sample orientations that will cover all the required peaks
without unnecessary redundancy.

@@

The goal of CrystalPlan was to make an easy-to-use software tool to aid users in
planning their experiments. By quickly making an experimental plan that will be
tailored to the needs of the scientist (enough peaks measured for a structure
analysis), but without spending time collecting redundant data, CrystalPlan will
greatly increase the scientific productivity and throughput of TOPAZ and other
similar beamlines. Users will be able to measure samples highly efficiently,
allowing them to explore parameters such as temperature and pressure more fully.

@@

\begin{itemize}
  \item Load a geometry file for a specific instrument (using a generic,
  flexible file format).
  \item Calculate the orientation of a crystal at different goniometer settings.
  \item Calculate the coverage of these orientations in reciprocal space,
  in volume and for single peaks.
  \item Take into account crystal Bravais lattice symmetries and the higher
  redundancy of higher space groups.
  \item Optimize the experiment plan automatically.
  \item Compare predicted and actual measurement of peaks.
\end{itemize}

@@

CrystalPlan can accomodate a variety of sample
orientation goniometers. A base Goniometer class (in the object-oriented
programming sense) is subclassed for specific types of goniometers, allowing a
unified interface. This base class acts as a \emph{universal goniometer},
which uses the conventional Eulerian goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from the base Goniometer can convert
the desired $\phi$, $\chi$, $\omega$ angles to the internal motor positions it requires. 
The class hierarchy allows for goniometer limitations: for example, TOPAZ is
currently equipped with a goniometer where $\chi$ is fixed at 45 degrees.
CrystalPlan handles this by limiting the sample orientation angles accessible to the user to $\phi$ and
$\omega$.

@@

Crystal lattice parameters (lengths $a$, $b$, $c$ and angles $\alpha$, $\beta$,
$\gamma$) are typed in or loaded from the ISAW UB matrix file. CrystalPlan then
generates a list of $hkl$ peaks with d-spacings larger than the $d_{min}$
specified. For each sample orientation being simulated and for each peak, a function calculates at
what wavelength $\lambda$ and in which direction it scatters. If $\lambda$ 
is within range of the limits of the instrument, the beam direction is projected
onto each detector face to determine where, on each detector surface, it will be
measured.

Each $hkl$ peak is a separate object that holds all predicted
recordings, including which detector measured it, at what (x,y) position on the
detector face, at what wavelength, and for which sample orientation. These data
are used to graphically display coverage and calculate statistics such as
redundancy.

@@

\subsubsection{Reciprocal Space Coverage Visualization.}
Figure \ref{fig:volume_view}a) shows a screenshot of the reciprocal space
coverage 3D interface. The 3D view can represent {\bf Q}-space in two ways: volume coverage, or single-peak
coverage. The view shown can be rotated, translated or zoomed using the mouse or
keyboard.

@@

Finally, both views display coverage statistics in the lower right
corner, showing the percentage of volume (or of the number of peaks) measured,
as well as the proportion of redundancy (voxels or peaks measured 2 or more
times).

\subsubsection{Single Reflection View.}
Figure \ref{fig:single_reflection} shows a screenshot of the GUI allowing one to
observe in detail the predicted measurements of a particular peak, by clicking
in the 3D view or by typing in the $hkl$ values. The position on the detector
and wavelength of measurement are displayed. An additional window (not shown) allows the user to place the peak on a particular
position on a detector by automatically finding the necessary sample orientation
angles. This is used, for example, to center important peaks or diffraction
directions in the most sensitive part of a detector.

@@

As part of the TOPAZ beamline commissioning process, a natrolite
(\ce{Na2Al2Si3O10} $\cdot$ \ce{2H2O}) single crystal was measured on a fixed
$45\degree\ \chi$ goniometer at room temperature and ambient atmosphere. An
initial short run (wavelength range from $\sim0.5$ \ang\ to $3.5$ \ang)
was processed using ISAW to determine the UB matrix and cell parameters of the crystal.

The resulting UB matrix was loaded into CrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations (Figure \ref{fig:natrolite_gui}a) that covered 97\% of the peaks,
taking into account the crystal Laue symmetry (orthorhombic or mmm) (Figure
\ref{fig:natrolite_gui}b). The eight orientations were run
unattended overnight.

Peaks were integrated using ISAW, and the resulting .integrate file was loaded
into CrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated peaks
found. Good agreement was found; the root-mean-square difference between the
predicted and actual measurement wavelength was only 0.016 \ang. The offsets in the peak
position were plotted, which helped show that the main contributor to this error
was a slight misalignment of the goniometer.

@@

@@

We have described CrystalPlan, a software program designed to quickly produce
efficient experiment plans for wide bandwidth time-of-flight Laue
diffractometers. This program will help increase scientific productivity at SNS
beamlines such as TOPAZ.

One planned future development for CrystalPlan include extending the software
to help predict coverage in inelastic neutron scattering experiments. Another
desired feature is the introduction of feedback loops to help determine the
right time to end data acquisition at a particular sample orientation.
CrystalPlan would interface with data analysis software such as ISAW and use
recently developed capabilities for real-time data analysis to monitor
the  signal to noise ratio on particular peaks. When the signal intensity
becomes sufficient, the sample will automatically be moved to the next sample
orientation, further adjusting data acquisition time for each specific sample.
Combined with advanced sample changers coming on-line at TOPAZ, this will allow
for extremely efficient, unattended and automated data acquisition.

@@

\begin{figure}
\caption{Sample screenshot showing the 3D view of single-crystal reflection
measurements. Each sphere represents an $hkl$ peak, color-coded to indicate the
number of measurements (dark blue: 0, cyan: 1, green: 2, yellow: 3, red: 4 or
more). a) without crystal symmetry, 2375 of 9481 reflections are measured, or
25\%. b) with crystal symmetry, 1276 of 1427 primary reflections are measured, or
89\%, and redundancy is also increased. }
\ifdograph
  \includegraphics{refl_nosym} 
\fi
a)
\ifdograph
  \includegraphics{refl_sym}
\fi
b)
\label{fig:refl_view}
\end{figure}

>>>>

\newif\ifdograph
\dographtrue % DO keep the graphs.
%\dographfalse % DONT put graphs - makes things faster

@@

\begin{abstract}
Beam time at large user program based x-ray and neutron scattering facilities is
in high demand and always at a premium. CrystalPlan, a highly efficient
experiment planning software has been developed to maximize the use of available
beamtime per sample per experiment. 
This program can calculate and optimize the data coverage of a crystal in
reciprocal space in a single-crystal diffraction time-of-flight experiment.
CrystalPlan can help a user build an experiment plan that will acquire the most
unique data possible, with sufficient coverage but limited redundancy, therefore
increasing scientific productivity. 
A user friendly GUI (Graphical User Interface) including a 3D viewer, an
automated coverage optimizer, and an option to reorient the crystal for the measurement 
of selected $hkl$s on specific detector positions 
are among its useful features. 
A sample use case of the program with the TOPAZ beamline at SNS will be
presented. 
\end{abstract}

@@

The TOPAZ beamline at the Spallation Neutron Source (SNS) at Oak Ridge National
Laboratory (ORNL), currently finishing commissioning, is a neutron
time-of-flight Laue diffractometer with high wavelength (time) resolution. It
is designed to acquire single-crystal diffraction data of small-to-moderate sized unit cells with  high
throughput. The final design calls for an array of 48 detectors covering three
quarters of $2\pi$ steradians. However as of this writing, only 14 of the 48
detectors are installed, giving only a quarter of $2\pi$ steradian coverage. 
To have sufficient data for a reliable structure solution and
refinement, an experimenter will typically want to measure $ > 85\%$ of the
unique reflections within a certain range of d-spacings. However, TOPAZ's wide
wavelength band combined with its complex geometry makes it difficult to come up with
an effective set of sample orientations that will cover all the required reflections
without unnecessary redundancy.

@@

The goal of CrystalPlan was to make an easy-to-use software tool to aid users in
planning their experiments. By quickly making an experimental plan that will be
tailored to the needs of the scientist (enough reflections measured for a structure
analysis), but without spending time collecting redundant data, CrystalPlan will
greatly increase the scientific productivity and throughput of TOPAZ and other
similar beamlines. Users will be able to measure samples highly efficiently,
allowing them to explore parameters such as temperature and pressure more fully.

@@

\begin{itemize}
  \item Load a geometry file for a specific instrument (using a generic,
  flexible file format).
  \item Calculate the orientation of a crystal at different goniometer settings.
  \item Calculate the coverage of these orientations in reciprocal space,
  in volume and for single reflections.
  \item Take into account crystal Bravais lattice symmetries and the higher
  redundancy of higher space groups.
  \item Optimize the experiment plan automatically.
  \item Compare predicted and actual measurement of reflections.
\end{itemize}

@@

CrystalPlan can accomodate a variety of sample
orientation goniometers. A base Goniometer class (in the object-oriented
programming sense) is subclassed for specific types of goniometers, allowing a
unified interface. This base class acts as a \emph{universal goniometer},
which uses the conventional Eulerian goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from the base Goniometer can convert
the desired $\phi$, $\chi$, $\omega$ angles to the internal motor positions it requires. 
The class hierarchy allows for goniometer limitations: for example, TOPAZ is
currently equipped with a goniometer where $\chi$ is fixed at 45 degrees
(at $\omega$ to the incident beam). CrystalPlan handles this by limiting the
sample orientation angles accessible to the user to $\phi$ and $\omega$.

@@

Crystal lattice parameters (lengths $a$, $b$, $c$ and angles $\alpha$, $\beta$,
$\gamma$) are typed in or loaded from the ISAW UB matrix file. CrystalPlan then
generates a list of $hkl$ reflections with d-spacings larger than the $d_{min}$
specified. For each sample orientation being simulated and for each reflection, a function calculates at
what wavelength $\lambda$ and in which direction it scatters. If $\lambda$ 
is within range of the limits of the instrument, the beam direction is projected
onto each detector face to determine where, on each detector surface, it will be
measured.

Each $hkl$ reflection is a separate object that holds all predicted
recordings, including which detector measured it, at what (x,y) position on the
detector face, at what wavelength, and for which sample orientation. These data
are used to graphically display coverage and calculate statistics such as
redundancy.

@@

\subsubsection{Reciprocal Space Coverage Visualization.}
Figure \ref{fig:volume_view}a) shows a screenshot of the reciprocal space
coverage 3D interface. The 3D view can represent {\bf Q}-space in two ways: volume coverage, or single-reflection
coverage. The view shown can be rotated, translated or zoomed using the mouse or
keyboard.

@@

Finally, both views display coverage statistics in the lower right
corner, showing the percentage of volume (or of the number of reflections) measured,
as well as the proportion of redundancy (voxels or reflections measured 2 or more
times).

\subsubsection{Single Reflection View.}
Figure \ref{fig:single_reflection} shows a screenshot of the GUI allowing one to
observe in detail the predicted measurements of a particular reflection, by clicking
in the 3D view or by typing in the $hkl$ values. The position on the detector
and wavelength of measurement are displayed. An additional window (not shown) allows the user to place the reflection on a particular
position on a detector by automatically finding the necessary sample orientation
angles. This is used, for example, to center important reflections or diffraction
directions in the most sensitive part of a detector.

@@

As part of the TOPAZ beamline commissioning process, a natrolite
(\ce{Na2Al2Si3O10} $\cdot$ \ce{2H2O}) single crystal was measured on a fixed
$45\degree\ \chi$ goniometer at room temperature and ambient atmosphere, with a
wavelength range from $\sim0.5$ \ang\ to $3.5$ \ang.
An initial short event file (each neutron detection event is saved with its
time-of-flight and position) was processed using ISAW to determine the UB
matrix and cell parameters of the crystal.

The resulting UB matrix was loaded into CrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations (Figure \ref{fig:natrolite_gui}a) that covered 97\% of the reflections,
taking into account the crystal Laue symmetry (orthorhombic or mmm) (Figure
\ref{fig:natrolite_gui}b). The eight orientations were run
unattended overnight.

Reflections were integrated using ISAW, and the resulting file was
loaded into CrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated reflections
found. Good agreement was found; the root-mean-square difference between the
predicted and actual measurement wavelength was only 0.016 \ang. The offsets in the reflection
position were plotted, which helped show that the main contributor to this error
was a slight misalignment of the goniometer.

@@

CrystalPlan was also successfully used in a similar way during experiments run
on the TOPAZ instrument by Yuri Janssen from State University of New York, Stony
Brook, and Henrik Clausen from University of Aarhus, Denmark.

@@

We have described CrystalPlan, a software program designed to quickly produce
efficient experiment plans for wide bandwidth time-of-flight Laue
diffractometers. This program will help increase scientific productivity at SNS
beamlines such as TOPAZ. With some relatively minor modifications, the program
could also be adapted for use at reactor sources. Monochromatic reactor sources
could be handled by adding support sweeps consisting of many motor positions;
Laue diffractometers would requiring detecting and removing reflections with chromatic overlap.

One planned future development for CrystalPlan include extending the software
to help predict coverage in inelastic neutron scattering experiments. Another
desired feature is the introduction of feedback loops to help determine the
right time to end data acquisition at a particular sample orientation.
CrystalPlan would interface with data analysis software such as ISAW and use
recently developed capabilities for real-time data analysis to monitor
the  signal to noise ratio on particular reflections. When the signal intensity
becomes sufficient, the sample will automatically be moved to the next sample
orientation, further adjusting data acquisition time for each specific sample.
Combined with advanced sample changers coming on-line at TOPAZ, this will allow
for extremely efficient, unattended and automated data acquisition.

@@

\begin{figure}
\caption{Sample screenshot showing the 3D view of single-crystal reflection
measurements. Each sphere represents an $hkl$ reflection, color-coded to indicate the
number of measurements (dark blue: 0, cyan: 1, green: 2, yellow: 3, red: 4 or
more). a) without crystal symmetry, 2375 of 9481 reflections are measured, or
25\%. b) with crystal symmetry, 1276 of 1427 primary reflections are measured, or
89\%, and redundancy is also increased. }
\ifdograph
  \includegraphics{refl_nosym} 
\fi
a)
\ifdograph
  \includegraphics{refl_sym}
\fi
b)
\label{fig:refl_view}
\end{figure}
